---
title: "How EDI Relates to randomizr, carat, coin, and Other Experiment-Design Packages"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

R has excellent packages for individual pieces of the randomized-experiment
workflow. This page maps EDI (Experimental Design and Inference) onto that
landscape honestly: what each neighboring package does well, when you should
prefer it, and what EDI does differently. It is written by EDI's author —
corrections from the other packages' perspectives are welcome on the
[issue tracker](https://github.com/kapelner/EDI/issues).

**EDI's one distinctive idea** is that a single design object carries the
assignment mechanism from randomization through analysis: you construct a
design (fixed or sequential), it assigns treatment, and the inference
classes then compute estimates, intervals, and tests *matched to that
mechanism* — including randomization tests that re-draw assignments from
the same algorithm that produced the data. Most neighboring packages cover
one stage of that pipeline; many of them do their stage with more options
than EDI does.

## At a glance

| Package | Stage it covers | Relative to EDI |
|---|---|---|
| `randomizr` | Assignment declaration (complete, blocked, clustered, stratified) | Assignment only; pairs with `estimatr`/`ri2` for analysis |
| `blockrand` | Permuted-block randomization lists for clinical trials | List generation only; no analysis |
| `Minirand` | Pocock-Simon minimization | Assignment only |
| `carat` | Covariate-adaptive randomization + some associated tests | Closest neighbor on the sequential-design side |
| `coin` | Permutation/conditional inference framework | Inference only; general-purpose, not design-coupled |
| `ri2` | Randomization inference (Gerber & Green) | Inference for declared designs; pairs with `randomizr` |
| `estimatr` | Design-based estimators with robust/cluster SEs | Estimation only; frequentist asymptotics |
| `pwr` | Closed-form power calculations | Analytic power; EDI's power is simulation-based |
| `simsurv` | Simulating survival data | Data generation only |
| `survival`, `icenReg` | General survival modeling (incl. interval censoring) | General-purpose modeling, not experiment-coupled |

## Assignment-stage packages

### randomizr (and its companions estimatr, ri2)

[`randomizr`](https://declaredesign.org/r/randomizr/) declares and draws
assignments — complete, simple, blocked, clustered, and blocked-clustered
randomization — with a clean, widely used API, and it belongs to the
DeclareDesign family where [`estimatr`](https://declaredesign.org/r/estimatr/)
supplies design-based estimators and [`ri2`](https://declaredesign.org/r/ri2/)
supplies randomization inference. That family's philosophy is close to
EDI's (the design should inform the analysis).

*Prefer that stack when* you want to declare a fixed design abstractly,
simulate over it with DeclareDesign, or you're already invested in that
ecosystem. *EDI differs in* bundling the pipeline into one object, adding
sequential/covariate-adaptive designs (which `randomizr` does not cover),
and providing per-response-type model-based estimators (GLMs, GLMMs,
survival, ordinal) alongside the design-based ones.

### blockrand and Minirand

[`blockrand`](https://cran.r-project.org/package=blockrand) generates
permuted-block randomization lists (varying block sizes, stratification) for
clinical trials; [`Minirand`](https://cran.r-project.org/package=Minirand)
implements Pocock-Simon minimization. Both are focused, dependable
generators of assignments — and both stop there.

*Prefer them when* you only need a randomization list to hand to a trial
coordinator. *EDI differs in* implementing the same schemes
(`DesignSeqOneByOneRandomBlockSize`, `DesignSeqOneByOnePocockSimon`) as live
design objects that subsequently drive analysis matched to the scheme —
e.g., randomization tests that replay minimization rather than assuming a
coin flip.

### carat

[`carat`](https://cran.r-project.org/package=carat) is the closest neighbor
on the sequential side: covariate-adaptive randomization procedures
(stratified biased coin, Hu-Hu, Pocock-Simon, and others) together with
some hypothesis tests derived for those procedures.

*Prefer carat when* you want specific covariate-adaptive procedures or
their dedicated asymptotic tests from that literature. *EDI differs in*
scope: the matching-on-the-fly (KK) family, matched-pair pooling strategies
(inverse-variance weighting and combined likelihood), six response types
with per-type estimator menus, resampling/exact inference throughout, and
the simulation framework.

## Inference-stage packages

### coin

[`coin`](https://cran.r-project.org/package=coin) is a mature, general
framework for permutation and conditional inference with a rich class of
test statistics. It permutes within strata you specify, but it is not
coupled to how treatment was actually assigned.

*Prefer coin when* you need a permutation test outside the
designed-experiment setting or a statistic EDI doesn't offer. *EDI differs
in* drawing its randomization distributions from the design's own
assignment algorithm (blocked, matched, minimized, rerandomized), which is
the difference between a permutation test and a *randomization* test when
the mechanism is not exchangeable-uniform.

## Power analysis

### pwr

[`pwr`](https://cran.r-project.org/package=pwr) computes closed-form power
for standard tests — instant and exact under its assumptions. *Prefer it
when* your planned analysis matches a textbook test. *EDI differs in*
computing power by Monte Carlo over the actual design x estimator you plan
to use (`SimulationFramework`), which is slower but answers the question
for analyses with no closed form (GLMMs, covariate-adaptive designs,
randomization tests), and additionally reports size and coverage
diagnostics.

```{r}
library(EDI)
# power of a matching-on-the-fly design analyzed by matched OLS,
# versus complete randomization analyzed by plain OLS:
sim = SimulationFramework$new(
  response_type = "continuous",
  design_classes_and_params = list(
    DesignSeqOneByOneKK21 = list(),
    DesignSeqOneByOneBernoulli = list()
  ),
  inference_classes_and_params = list(
    InferenceContinKKOLSIVWC = list(),
    InferenceContinOLS = list()
  ),
  n = 100, p = 3, Nrep_W = 1000L, betaT = 0.5,
  results_filename = "power_comparison.csv.bz2",
  continue_from_last_result_row = FALSE
)
sim$run()
SimulationFrameworkReport$new(sim)$summarize()
```

## Survival modeling and simulation

### survival, icenReg, simsurv

[`survival`](https://cran.r-project.org/package=survival) is R's canonical
survival toolbox and [`icenReg`](https://cran.r-project.org/package=icenReg)
specializes in interval-censored regression; both are general-purpose.
[`simsurv`](https://cran.r-project.org/package=simsurv) simulates survival
data flexibly.

*Prefer them for* observational survival analysis, model families EDI
lacks, or standalone data simulation. *EDI differs in* embedding survival
inference in the experiment pipeline — Cox, Weibull AFT with exact / left /
right / interval censoring in one `y`/`y_L`/`y_R` likelihood, RMST,
log-rank/Gehan with randomization p-values, and matched-pair frailty models
— and its survival simulations run inside the same power framework as every
other response type. (EDI itself Imports `survival`.)

## Using them together

These are complements more often than competitors: generate data with
`simsurv` and analyze the designed part with EDI; sanity-check an EDI
randomization p-value against `coin` on an exchangeable design (they should
agree there); use `pwr` for a quick analytic bound before committing to a
long simulation. Where a neighboring package is the better tool for your
problem, use it — the point of this page is to make that call easy to get
right.

```{r}
# EDI's end-to-end signature: one object from design to inference
library(EDI)
des = DesignSeqOneByOneKK21$new(n = 40, response_type = "continuous")
for (i in 1 : 40) {
  des$add_one_subject_to_experiment_and_assign(X[i, , drop = FALSE])
}
des$add_all_subject_responses(y)
InferenceSuite$new(des)$run_all_inference(screen = TRUE)$results_table
```
