---
title: "Try EDI in Your Browser (webR) — No Installation Needed"
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

You can run EDI (Experimental Design and Inference) entirely in your web
browser — no R installation, no compiler, nothing downloaded to your
machine beyond the page itself. This works because
[webR](https://docs.r-wasm.org/webr/latest/) compiles R itself to
WebAssembly, and EDI publishes WebAssembly binaries (of the same C++
kernels the native package uses) through
[R-universe](https://kapelner.r-universe.dev). Every one of EDI's hard
dependencies also has a WebAssembly build, so the whole stack loads in the
browser.

## Step 1: open the webR REPL

Open **<https://webr.r-wasm.org/latest/>** in a new tab. After a few
seconds you get a working R console running locally in your browser —
nothing you type leaves your machine.

## Step 2: install EDI

Paste this into the webR console (the download is a few tens of MB the
first time; give it a minute):

```{r}
webr::install("EDI",
  repos = c("https://kapelner.r-universe.dev", "https://repo.r-wasm.org"))
library(EDI)
```

## Step 3: run a complete experiment

Design, assign, respond, infer — the full EDI workflow, sized to run
comfortably in a browser tab:

```{r}
library(EDI)

n = 40
X = data.frame(age = rnorm(n, 60, 8), weight = rnorm(n, 80, 12))

# rerandomization: re-draw the assignment until covariate balance passes
des = DesignFixedRerandomization$new(n = n, response_type = "continuous")
des$add_all_subjects_to_experiment(X)
des$assign_w_to_all_subjects()
y = rnorm(n) + 0.5 * des$get_w()          # (simulated outcomes for the demo)
des$add_all_subject_responses(ys = y)

inf = InferenceContinOLS$new(des)
inf$compute_estimate()                    # covariate-adjusted treatment effect
inf$compute_asymp_confidence_interval()   # asymptotic 95% CI
inf$compute_rand_two_sided_pval()         # exact randomization (design-based) test
```

Or a sequential matching-on-the-fly trial, analyzed by every applicable
procedure at once:

```{r}
seq_des = DesignSeqOneByOneKK21$new(n = 20, response_type = "continuous")
for (i in 1 : 20) {
  # each subject is matched and assigned on arrival
  seq_des$add_one_subject_to_experiment_and_assign(data.frame(x = rnorm(1)))
}
seq_des$add_all_subject_responses(rnorm(20))

suite = InferenceSuite$new(seq_des)
res = suite$run_all_inference(screen = TRUE)
res$results_table
```

## What to expect (honest caveats)

The browser build is for *trying* EDI, not for production analyses:

- **It is single-threaded.** WebAssembly R has no OpenMP, so the
  parallelized bootstrap/randomization loops run serially.
- **It is a generic build.** The wasm binaries carry none of the
  machine-specific tuning (`-march=native`, machine-calibrated dispatch
  policies) that makes native EDI fast. Expect resampling-heavy calls to be
  one to two orders of magnitude slower than a tuned native install.
- **Memory lives in the tab.** Very large simulations can exhaust the
  browser's WebAssembly memory; keep `n` and replication counts modest.

When you're ready to use EDI for real work, install it natively — and
compile from source so the build tunes itself to your CPU:

```{r}
install.packages("EDI", type = "source")   # builds with -march=native
tune_EDI_for_this_machine()                # calibrates dispatch policies
```

(Prebuilt native binaries, no toolchain needed, are also available:
`install.packages("EDI", repos = c("https://kapelner.r-universe.dev",
"https://cloud.r-project.org"))`.)
